Layered Formal Verification of an Industrial TCP/IP Stack

### 2021-05-28 SecDev21/paper.tex
slightly change the part on symbolic execution

Generally, every TCP communication session will go through the following three phases (if no error occurs): 1)~opening the connection, 2)~sending and receiving the data, and 3)~closing the connection. The behavior of the three-phase communication can be described by the state machine given in \cref{Fig:statemachine}. An edge represents a state transition.
An edge-label in the format of $\frac{\ \ x \ \ }{\ \ y \ \ }$ represents the actions associated to the state transition, where $x$ is triggering the transition, and $y$ is produced by the transition.
The trigger $x$ is either an explicit action like \textit{Close}, or the arrival of specific flags, while $y$ always represents the flags transmitted in response to the action that caused the state transition, when present.
An explicit action is either a user-triggered action (through the user's API) or an action automatically triggered by a timeout event. The flags are embedded in the header section of a transmitted segment and can be one of the following:
\begin{itemize}
\item \ack: the last message received by the sender is acknowledged.
\item \syn: this flag is sent to establish a connection.
\item \fin: this flag is sent to close the connection.
\item \rst: this flag is sent to reset the connection, often when an error
occurred on one or the other side.
\end{itemize}

The state machine in \cref{Fig:statemachine} does not represent the complete protocol specification; for example, it does not reflect error conditions or any actions which are not connected with the state changes. Instead, it gives an overview of all the possible states a TCP connection could reach over its lifetime. Each state has a meaning. \slisten represents waiting for a connection request from any remote TCP and port. \ssynsent and \ssynrcv represent the initialization of the connection that ends when \sestab is reached. \sfwone, \sfwtwo, \sclosing, and \stimewait are a group of states that represent waiting for a connection termination request from the remote TCP. \sclosew and \slastack represent waiting for a connection termination request from the local user. Finally, \sclosed represents no connection state at all.

@@

The original \emph{CycloneTCP} library provides no guarantees that the implementation respects the TCP functional specification. Thus, there is no assurance that the implementation will only allow valid transitions between the different states that a TCP session can exhibit. We aim to use the SPARK technology to verify that the implementation respects the TCP automaton, given in \cref{Fig:statemachine}. Although, this work is focused on hardening the API of the user task, mainly the high-level user functions, to verify the state transitions the rest of the TCP protocol has to be taken into account. The reason for this is two-fold. First, although user-functions can trigger state transitions, the actual transitions are done by library functions that do not belong to the user-task set of functions. Second, other parts of the library can trigger state transitions during and between the user-task function calls, which can affect the intermediate and final states that a user related function can exhibit.

@@

To model any possible state-changes upon the reception of a segment in SPARK, the function \spark{TCP_Process_One_Segment} is introduced.
The behavior of \spark{TCP_Wait_For_Events} is modeled by the ghost function \spark{TCP_Wait_For_Events_Proof} shown in \cref{algo:waitForEvents}. It computes the set of possible final states after the completion of an event by reusing the function \spark{TCP_Process_One_Segment}. Both functions have the same behavior, and thus, the contract of the function \spark{TCP_Wait_For_Events_Proof} is also attached to \spark{TCP_Wait_For_Events}. This ghost function is proved automatically without even a loop invariant thanks to loop unrolling in GNATprove.

@@

Rewriting the part that processes the incoming segments was out of the scope of
this work. Instead, we have opted for the use of symbolic execution on the
original CycloneTCP C code.
Symbolic execution is a mean to execute programs with symbolic values rather
than concrete values. Symbolic execution has already been successfully
used with other verification methods.
Indeed, in~\cite{vanoverberghe2008using} the authors present how to use symbolic
execution to improve deductive verification, and in~\cite{kassios2012comparing}
the authors explain that symbolic execution can be good at inferring contracts.
For our work, KLEE has been used. It is built on top of LLVM as it takes LLVM
bytecode in input and works well with C code thanks to a very simple interface.

@@

The first step in order to extract contracts was to read the state machine to
know all the possible transitions that can be done by the reception of a flag,
and adapt it to
the subtleties of CycloneTCP. The next step is to write a driver to run KLEE and
know if the contracts that we have inferred are correct or not.
A simplified example is shown in figure~\ref{code:kleedriver}. The first lines
create a symbolic incoming segment to process as well as a symbolic socket.
At line~\ref{code:kleedriver:tcpProcessSegment} the function that processes the
segment is called. It is followed at line~\ref{code:kleedriver:assert} by a KLEE
builtin that checks if an assert is valid for all the symbolic paths
explored before.

Once the contract is validated by KLEE, it can be copied out on the SPARK
function \spark{TCP_Process_One_Segment}. With this step done, the concurrency
challenge has been solved and the function
\spark{TCP_Wait_For_Events} have a correct and an accurate contract, which
makes possible the verification of the functional correctness of the TCP user
functions thanks to SPARK.

>>>>

Generally, every TCP communication session will go through the following three phases (if no error occurs): 1)~opening the connection, 2)~sending and receiving the data, and 3)~closing the connection. The behavior of the three-phase communication can be described by the state machine given in \Cref{Fig:statemachine}. An edge represents a state transition.
An edge-label in the format of $\frac{\ \ x \ \ }{\ \ y \ \ }$ represents the actions associated to the state transition, where $x$ is triggering the transition, and $y$ is produced by the transition.
The trigger $x$ is either an explicit action like \textit{Close}, or the arrival of specific flags, while $y$ always represents the flags transmitted in response to the action that caused the state transition, when present.
An explicit action is either a user-triggered action (through the user's API) or an action automatically triggered by a timeout event. The flags are embedded in the header section of a transmitted segment and can be one of the following:
\begin{itemize}
\item \ack: the last message received by the sender is acknowledged.
\item \syn: this flag is sent to establish a connection.
\item \fin: this flag is sent to close the connection.
\item \rst: this flag is sent to reset the connection, often when an error
occurred on one or the other side.
\end{itemize}

The state machine in \Cref{Fig:statemachine} does not represent the complete protocol specification; for example, it does not reflect error conditions or any actions which are not connected with the state changes. Instead, it gives an overview of all the possible states a TCP connection could reach over its lifetime. Each state has a meaning. \slisten represents waiting for a connection request from any remote TCP and port. \ssynsent and \ssynrcv represent the initialization of the connection that ends when \sestab is reached. \sfwone, \sfwtwo, \sclosing, and \stimewait are a group of states that represent waiting for a connection termination request from the remote TCP. \sclosew and \slastack represent waiting for a connection termination request from the local user. Finally, \sclosed represents no connection state at all.

@@

The original \emph{CycloneTCP} library provides no guarantees that the implementation respects the TCP functional specification. Thus, there is no assurance that the implementation will only allow valid transitions between the different states that a TCP session can exhibit. We aim to use the SPARK technology to verify that the implementation respects the TCP automaton, given in \Cref{Fig:statemachine}. Although, this work is focused on hardening the API of the user task, mainly the high-level user functions, to verify the state transitions the rest of the TCP protocol has to be taken into account. The reason for this is two-fold. First, although user-functions can trigger state transitions, the actual transitions are done by library functions that do not belong to the user-task set of functions. Second, other parts of the library can trigger state transitions during and between the user-task function calls, which can affect the intermediate and final states that a user related function can exhibit.

@@

To model any possible state-changes upon the reception of a segment in SPARK, the function \spark{TCP_Process_One_Segment} is introduced.
The behavior of \spark{TCP_Wait_For_Events} is modeled by the ghost function \spark{TCP_Wait_For_Events_Proof} shown in \Cref{algo:waitForEvents}. It computes the set of possible final states after the completion of an event by reusing the function \spark{TCP_Process_One_Segment}. Both functions have the same behavior, and thus, the contract of the function \spark{TCP_Wait_For_Events_Proof} is also attached to \spark{TCP_Wait_For_Events}. This ghost function is proved automatically without even a loop invariant thanks to loop unrolling in GNATprove.

@@

The soundness of our verificatin requires all the contracts to be correct, in particular this is true for the contract of \spark{TCP_Process_One_Segment}.
Get this contract can be done by rewriting in SPARK the part that processes the
incoming segments, but this is out of the scope of this work.
Instead, we have opted for the use of symbolic execution on the
original CycloneTCP C code.
Symbolic execution is a mean to execute programs with symbolic values rather
than concrete values. Symbolic execution has already been successfully
used with other verification methods.
Indeed, in~\cite{vanoverberghe2008using} the authors present how to use symbolic
execution to improve deductive verification, and in~\cite{kassios2012comparing}
the authors explain that symbolic execution can be good at inferring contracts.
For our work, KLEE\footnote{\url{https://klee.github.io}} has been used.
It is built on top of LLVM as it takes LLVM
bytecode in input and works well with C code thanks to a very simple interface.

@@

The first step in order to extract contracts was to read the state machine to
know all the possible transitions that can be done by the reception of a flag,
and adapt it to
the subtleties of CycloneTCP. The next step is to write a driver to run KLEE and
know if the contracts that we have inferred are correct or not.
A simplified example is shown in \Cref{code:kleedriver}. The first lines
create a symbolic incoming segment to process and a symbolic socket.
At \cref{code:kleedriver:tcpProcessSegment} the function that processes the
segment is executed. It is followed at \cref{code:kleedriver:assert} by a KLEE
builtin that checks if the code leads to a valid state identified in the
protocol’s specification for all the paths explored.

Once the contract is validated by KLEE, it can be copied out on the SPARK
function \spark{TCP_Process_One_Segment}. With this step done, the concurrency
challenge has been solved and the function
\spark{TCP_Wait_For_Events} has a correct and an accurate contract, which
makes possible the verification of the functional correctness of the TCP user
functions thanks to SPARK.
